Grappa: Scalable Computing on Mass-Market Hardware for Irregular, Low-Locality Applications

### 2012-07-20 doc/asplos13/grappa.tex
pass on section 3

\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.5\columnwidth]{figs/grappa-crappy}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:grappa} Grappa runtime system \vspace{-4ex}}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

In this section we describe the Grappa runtime system
(Figure~\ref{fig:grappa}). Grappa is a software runtime system that
allows a commodity distributed-memory HPC cluster to be programmed as
if it were a single large shared-memory machine. Grappa is designed to
smooth over some of the performance discontinuities in commodity
hardware, giving good performance when there is little locality to be
exploited while allowing the programmer to exploit it when it is
available.

Grappa provides three key capabilities. The first is a distributed
shared \textbf{global memory} with high aggregate random access
bandwidth. The second is a \textbf{tasking} system with lightweight
multithreading and automatic load balancing through work
stealing. Finally, under the hood, a \textbf{communication} layer uses
network packet scheduling and aggregation to achieve high performance
while remaining largely invisible to applications.

In the following sections, we discuss the programmer's view of
Grappa's main capabilities, including a subset of the Grappa API. In
section~\ref{implementation}, we discuss the details of their
implementation.

@@

The basic unit of execution in Grappa is the {\em task}. Tasks are
small; just a function pointer and some arguments. Tasks are only
queued when they are spawned; later, when resources are free, they are
allocated a stack, bound to a core, and executed.

@@

All Grappa scheduling is done in userspace, and minimal state is
saved, in order to minimize the context switch time. In our
experiments, we saw context switch times as low as \TODO{40ns}.

This lightweight context switching is Grappa's key enabling
feature. It gives us the ability to tolerate latency. Given that
ability, we are able to trade latency for throughput: by {\em
  increasing} latency in key components in the system we are able to
increase our aggregate random access bandwidth, our synchronization
bandwidth, and our ability to tolerate load.

@@

A programmer's goal in coding with Grappa should be to express as much
parallelism as possible without worrying about where it will execute.
Grappa then chooses where and when to execute this parallelism,
scheduling as much work as necessary to tolerate network latencies and
keep the systems' cores busy.

@@

\subsection{Accessing memory}

@@

Large data that is expected to be accessed with low locality is stored in Grappa's global memory. All
global data must be accessed through calls into Grappa's API, shown in
Figure~\ref{fig:accessing-memory}.

Grappa provides two methods for {\emph storing} data in the global memory. The
first is a distributed heap striped across all the machines in the
system. The \texttt{global\_malloc} and \texttt{global\_free} calls
are used to allocate and deallocate memory in the global heap; on
allocation, a global pointer is returned. Grappa also allows any local
data on a core's stacks or heap to be exported to the global address
space to be made accessible to other cores across the system.

There are also two approaches to {\emph accessing} global memory. When the
programmer expects a computation on shared data to have spatial
locality to exploit, {\em cache} operations may be used. When there is no locality to
exploit, {\em delegate} operations are used. Since these operations
are expected communicate with other nodes and have high latency,
\TODO{...}

With a cache, applications can instruct Grappa to fetch a global
pointer of any length and return a local pointer to a cached copy of
the global memory. Under the hood, Grappa performs the mechanics of
gathering chunks of data from multiple system nodes and presenting a
conventional appearing linear block of memory as a pointer into a
cache. Grappa cache operations have the usual read-only and read-write
variants, along with a write-only variant used to initialize data
structures. Cache operations exploit spatial locality by reducing the number of small network
messages accessing contguous data. Languages for distributed
shared memory systems have done optimizations to achieve a similar goal. The UPC compiler
coalesces struct and array accesses into 
remote get/put \cite{Chen:2005}, or Fortran D compiler's message
vectorization that hoists small messages out of a loop
\cite{FortranD:1992}. Caching in Grappa additionally provides a
mechanism for exploiting temporal locality by operating on the data locally.

When the access pattern has low-locality, it is more efficient
to modify the data on its home core rather than bringing a copy to the
requesting core and returning it after modification. Delegate
operations provide this capability. Applications can dispatch
computation to be performed on individual machine-word sized chunks of
global memory to the memory system itself (e.g.,
\emph{fetch-and-add}).

@@

\subsection{Under the hood}

@@

The first is a dynamic load balancing mechanism that uses work stealing to
move tasks from busy nodes to idle nodes. When a core detects that it
has idle execution resources and no new tasks to run, it issues steal
requests over the network. The core continues to execute remaining
active work while the steal requests are in flight. When a steal
request succeeds, it returns with more tasks that can be matched with
execution resources and scheduled.

The second is a network packet scheduling and aggregation system. From
the perspective of the application, all communication between tasks in
Grappa takes place either through global shared memory or remote task
spawns. Both of these requests are ultimately translated into explicit
inter-system-node/core communication requests that are dispatched to
the \gasnet~library. Most of \checkme{our requests} are small, and commodity
networks do not transport small packets efficiently. In order to make
more efficient use of bandwidth, Grappa aggregates packets going to the
same destination until it can form a larger packet that will make more
efficient use of the network. In this way, Grappa trades per-request
latency for application-level throughput.

>>>>

\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.5\columnwidth]{figs/grappa-crappy}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:grappa} Grappa runtime system \vspace{-4ex}\TODO{should we use the canonical Grappa picture here agumented withe the software stack?}}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

In this section we describe the Grappa runtime system
(Figure~\ref{fig:grappa}). 
%%% Moved to intro
% Grappa is a software runtime system that allows a commodity distributed-memory
% HPC cluster to be programmed as if it were a single large shared-memory
% machine.
% Grappa is designed to
% smooth over some of the performance discontinuities in commodity
% hardware, giving good performance when there is little locality to be
% exploited while allowing the programmer to exploit it when it is
% available.
Grappa provides three key capabilities. The first is a distributed
shared \textbf{global memory} with high aggregate random access
bandwidth. The second is a \textbf{tasking} system with lightweight
multithreading and automatic load balancing through work
stealing. Finally, under the hood, a \textbf{communication} layer uses
network packet scheduling and aggregation to achieve high performance
while remaining largely invisible to applications. We now discuss the programmer's view of
Grappa's main capabilities, including a subset of the Grappa API. In
Section~\ref{implementation}, we discuss the details of their
implementation.

\TODO{Should we have an example of code here, and walk thru as we explain the
components and parts of the API?}

@@

The basic unit of execution in Grappa is the {\em task}. Tasks are
small; just a function pointer and arguments. Tasks are only
queued when they are spawned; later, when resources are free, they are
allocated a stack, bound to a core, and executed.

@@

All Grappa scheduling is done in userspace, and minimal state is
saved, in order to minimize the context switch time. In our
experiments, we saw context switch times as low as \TODO{40ns}.
This lightweight context switching is Grappa's key enabling
feature. It gives us the ability to tolerate latency. Given that
ability, we are able to trade latency for throughput: by {\em
  increasing} latency in key components in the system we are able to
increase our aggregate random access bandwidth, our synchronization
bandwidth, and our ability to tolerate load.

@@

A programmer's goal in coding with Grappa should be to express as much
parallelism as possible without worrying about where it will execute.
Grappa then chooses where and when to exploit this parallelism,
scheduling as much work as necessary to tolerate network latencies and
keep cores busy.

@@

\subsection{Memory}

@@

Large data that is expected to be shared and accessed with low locality is stored in Grappa's global memory. All
global data must be accessed through calls into Grappa's API, shown in
Figure~\ref{fig:accessing-memory}.

\paragraph{Global memory allocation}
Grappa provides two methods for {\emph storing} data in the global memory. The
first is a distributed heap striped across all the machines in the
system. The \texttt{global\_malloc} and \texttt{global\_free} calls
are used to allocate and deallocate memory in the global heap; on
allocation, a global pointer is returned. Grappa also allows any local
data on a core's stacks or heap to be exported to the global address
space to be made accessible to other cores across the system.

\paragraph{Global memory access}
There are two approaches to {\emph accessing} global memory. When the
programmer expects a computation on shared data to have spatial
locality to exploit, {\em cache} operations may be used. When there is no locality to
exploit, {\em delegate} operations are used. Since these operations
are expected communicate with other nodes and have high latency,
\TODO{the runtime systems might context switch? what do we want to say here?}

\paragraph{Explicit caching}
With Grappa's explicit caching support, applications can instruct Grappa to fetch a global
pointer of any length and return a local pointer to a cached copy of
the global memory. Under the hood, Grappa performs the mechanics of
gathering chunks of data from multiple system nodes and presenting a
conventional appearing linear block of memory as a pointer into a
cache. Grappa cache operations have the usual read-only and read-write
variants, along with a write-only variant used to initialize data
structures. Cache operations exploit spatial locality by reducing the number of small network
messages accessing contiguous data. Languages for distributed
shared memory systems have done optimizations to achieve a similar goal. The UPC compiler
coalesces struct and array accesses into 
remote get/put \cite{Chen:2005}, or Fortran D compiler's message
vectorization that hoists small messages out of a loop
\cite{FortranD:1992}. Caching in Grappa additionally provides a
mechanism for exploiting temporal locality by operating on the data locally.

\paragraph{Delegate operations}
When the access pattern has low-locality, it is more efficient
to modify the data on its home core rather than bringing a copy to the
requesting core and returning it after modification. Delegate
operations provide this capability. Applications can dispatch
computation to be performed on individual machine-word sized chunks of
global memory to the memory system itself (e.g.,
\emph{fetch-and-add}). \TODO{should we bring up delegation in the intro?}

@@

\subsection{Load Balancing and Network Aggregation}

@@

The first is a distributed dynamic load balancing mechanism that uses work stealing to
move tasks from busy nodes to idle nodes. When a core detects that it
has idle execution resources and no new tasks to run, it issues steal
requests over the network. The core continues to execute remaining
active work while the steal requests are in flight. When a steal
request succeeds, it returns with more tasks that can be matched with
execution resources and scheduled.

The second is a network packet scheduling and aggregation system. From
the perspective of the application, all communication between tasks in
Grappa takes place either through global shared memory or remote task
spawns. Both of these requests are ultimately translated into explicit
inter-system-node/core communication requests that are dispatched to
the \gasnet~library. Most of \checkme{our requests} are small, and commodity
networks do not transport small packets efficiently. In order to make
more efficient use of bandwidth, Grappa aggregates packets going to the
same destination until it can form a larger packet that will make more
efficient use of the network. In this way, Grappa trades per-request
latency for application-level throughput. \TODO{Jacob, can you add more details here on how this works?}

### 2012-07-20 doc/asplos13/intro.tex
pass on section 3

We have developed Grappa:  a latency tolerant runtime for scaling irregular applications on x86 clusters.  Grappa leverages as much freely available and commodity infrastructure as possible. We use unmodified Linux for the operating system.  An off the shelf user-mode infiniband device driver stack~\cite{Melonox?} is used for network interfacing.  MPI~\cite{mpi} is used for process setup and tear down.  GASnet~\cite{gasnet} is used as the underlying mechanism for remote memory reads and writes using active message invocations. To this commodity hardware and software mix Grappa adds three main software components: (1) a lightweight tasking layer that supports a context switch in as few as \checkme{40ns} and distributed load balancing; (2) a distributed shared memory layer that supports customary operations such as read and write as well as remote operations such as fetch-and-add~\cite{fetchandadd}; and (3) a message aggregation layer that combines short messages to mitigate the aforementioned problem that commodity networks are designed to achieve peak bandwidth only on large packet sizes, yet irregular applications tend to fetch only a handful of bytes at a time.  Each of these components will be described in more detail in Section~\ref{sec:grappa}.

>>>>

In this paper we introduce Grappa, a software runtime system that allows a
commodity x86 distributed-memory HPC cluster to be programmed as if it were a
single large shared-memory machine and provides scalable performance for
irregular applications. Grappa is designed to smooth over some of the
performance discontinuities in commodity hardware, giving good performance
when there is little locality to be exploited while allowing the programmer to
exploit it when it is available. Grappa leverages as much freely available and
commodity infrastructure as possible. We use unmodified Linux for the
operating system. An off the shelf user-mode infiniband device driver
stack~\cite{Melonox?} is used for network interfacing. MPI~\cite{mpi} is used
for process setup and tear down. GASnet~\cite{gasnet} is used as the
underlying mechanism for remote memory reads and writes using active message
invocations. To this commodity hardware and software mix Grappa adds three
main software components: (1) a lightweight tasking layer that supports a
context switch in as few as \checkme{40ns} and distributed load balancing; (2)
a distributed shared memory layer that supports customary operations such as
read and write as well as remote operations such as
fetch-and-add~\cite{fetchandadd}; and (3) a message aggregation layer that
combines short messages to mitigate the aforementioned problem that commodity
networks are designed to achieve peak bandwidth only on large packet sizes,
yet irregular applications tend to fetch only a handful of bytes at a time.

Each of these components will be described in more detail in Section~\ref{sec:grappa}.
